---
title: "Lab 3 MATH 342W"
author: "Ye Htut Maung"
output: pdf_document
date: "11:59PM February 16"
---



## Perceptron

You will code the "perceptron learning algorithm" for arbitrary number of features p. Take a look at the comments above the function. Respect the spec below:

```{r}
#' TO-DO: Provide a name for this function 
#' Perceptron Learning Algorithm
#'
#' TO-DO: Explain what this function does in a few sentences
#' This function implements the Perceptron Learning Algorithm to find a separating hyperplane for binary classification.  
#' If the data is perfectly linearly separable, the algorithm converges to a weight vector `w` that classifies all points correctly.  
#' Otherwise, it runs for `MAX_ITER` iterations and returns the best weight vector found.
#'
#' @param Xinput      TO-DO: Explain this -The training data features as an n x p matrix.
#' @param y_binary    TO-DO: Explain this - The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param MAX_ITER    TO-DO: Explain this - maximum iteration that will run if the algorithm can't find w with no error
#' @param w           TO-DO: Explain this - initial weight vector that has p + 1 length
#'
#' @return            The computed final parameter (weight) as a vector of length p + 1
perceptron_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 1000, w = NULL){
  #TO-DO
  p = ncol(Xinput)
  n = nrow(Xinput)
  if (is.null(w)) {
    w = rep(0,time = p + 1)
  }
  X = cbind(1, Xinput)
  for (t in 1:MAX_ITER) {
    
   w_previous = w
   y_hat =  as.numeric(X %*% w > 0)
   for (i in 1:n) {
     w = w + X[i, ] * (y_binary[i] - y_hat[i])
   }
   
   if (identical(w, w_previous)){
     break
   }
  }
  w
}
```

To understand what the algorithm is doing - linear "discrimination" between two response categories, we can draw a picture. First let's make up some very simple training data D.

```{r}
Xy_simple = data.frame(
 response = factor(c(0, 0, 0, 1, 1, 1)), #nominal
 first_feature = c(1, 1, 2, 3, 3, 4),    #continuous
 second_feature = c(1, 2, 1, 3, 4, 3)    #continuous
)
```

We haven't spoken about visualization yet, but it is important we do some of it now. Thus, I will write this code for you and you will just run it. First we load the visualization library we're going to use:

```{r}
pacman::p_load(ggplot2)
```

We are going to just get some plots and not talk about the code to generate them as we will have a whole unit on visualization using `ggplot2` in the future.

Let's first plot y by the two features so the coordinate plane will be the two features and we use different colors to represent the third dimension, y.

```{r}
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response)) + 
  geom_point(size = 5)
simple_viz_obj
```

TO-DO: Explain this picture.
Red represents 0 and blue represent 1 and on the picture it is linearly separable data.

Now, let us run the algorithm and see what happens:

```{r}
w_vec_simple_per = perceptron_learning_algorithm(
  cbind(Xy_simple$first_feature, Xy_simple$second_feature), as.numeric(Xy_simple$response == 1))
w_vec_simple_per
```

Explain this output. What do the numbers mean? What is the intercept of this line and the slope? You will have to do some algebra.

TO-DO
The output are the weight vector. w consists of (w0, w1, w2) so (-17, 4, 4). when I substitute to the equation, it becomes -17 + 4x1+ 4x2 = 0
After calculation, I get x2 = (17/4) - x1. so the intercept is 17/4 or 4.25 and the slope is -1.


```{r}
simple_perceptron_line = geom_abline(
    intercept = -w_vec_simple_per[1] / w_vec_simple_per[3], 
    slope = -w_vec_simple_per[2] / w_vec_simple_per[3], 
    color = "orange")
simple_viz_obj + simple_perceptron_line
```

Explain this picture. Why is this line of separation not "satisfying" to you?

TO-DO 
The line of the separation is not satisfying to me because it is not in the middle of two supporting vectors or it is not in the middle of the empty space.


For extra credit, program the maximum-margin hyperplane perceptron that provides the best linear discrimination model for linearly separable data. Make sure you provide ROxygen documentation for this function.

```{r}
#TO-DO
```




## Support Vector Machine vs. Perceptron

We recreate the data from the previous lab and visualize it:

```{r}
pacman::p_load(ggplot2)
Xy_simple = data.frame(
 response = factor(c(0, 0, 0, 1, 1, 1)), #nominal
 first_feature = c(1, 1, 2, 3, 3, 4),    #continuous
 second_feature = c(1, 2, 1, 3, 4, 3)    #continuous
)
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response)) + 
  geom_point(size = 5)
simple_viz_obj
```

Use the `e1071` package to fit an SVM model to the simple data. Use a formula to create the model, pass in the data frame, set kernel to be `linear` for the linear SVM and don't scale the covariates. Call the model object `svm_model`. Otherwise the remaining code won't work.

```{r}
pacman::p_load(e1071)
svm_model = svm(
  formula = response ~ ., 
  data = Xy_simple, 
  kernel = "linear", 
  scale = FALSE
)
```

and then use the following code to visualize the line in purple:

```{r}
w_vec_simple_svm = c(
  svm_model$rho, #the b term
  -t(svm_model$coefs) %*% cbind(Xy_simple$first_feature, Xy_simple$second_feature)[svm_model$index, ] # the other terms
)
simple_svm_line = geom_abline(
    intercept = -w_vec_simple_svm[1] / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "purple")

simple_viz_obj + simple_svm_line
```

Source the `perceptron_learning_algorithm` function from lab 2. Then run the following to fit the perceptron and plot its line in orange with the SVM's line:

```{r}
w_vec_simple_per = perceptron_learning_algorithm(
  cbind(Xy_simple$first_feature, Xy_simple$second_feature),
  as.numeric(Xy_simple$response == 1)
)
simple_perceptron_line = geom_abline(
    intercept = -w_vec_simple_per[1] / w_vec_simple_per[3], 
    slope = -w_vec_simple_per[2] / w_vec_simple_per[3], 
    color = "orange")

simple_viz_obj + simple_perceptron_line + simple_svm_line
```

Is this SVM line a better fit than the perceptron?

TO-DO
Yes. This SVM line is a better fit than the perceptron as it is in more middle than the perceptron line.

Now write your own implementation of the linear support vector machine algorithm using the Vapnik objective function we discussed in class.

```{r}
pacman::p_load(optimx)
#' This function implements the hinge-loss + maximum margin linear support vector machine algorithm of Vladimir Vapnik (1963).
#'
#' @param Xinput      The training data features as an n x p matrix.
#' @param y_binary    The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param MAX_ITER    The maximum number of iterations the algorithm performs. Defaults to 5000.
#' @param lambda      A scalar hyperparameter trading off margin of the hyperplane versus average hinge loss.
#'                    The default value is 1.
#' @return            The computed final parameter (weight) as a vector of length p + 1
#' check mate package
linear_svm_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 5000, lambda = 1){
  #TO-DO
  p = ncol(Xinput)
  n = nrow(Xinput)
  
  w = rep(0,time = p + 1)
  X = cbind(1, Xinput)
  objective_function = function(w) {
    return (mean(pmax(0, (y_binary - (1/2)) * ( (X %*% w) - 1/2))) + lambda * sqrt(sum(w^2)))
  }
  opt_obj = optimx(w, objective_function, method = "BFGS")
  return (as.numeric(opt_obj$par))

}

```
``

If you wrote code (the extra credit), run your function using the defaults and plot it in brown vis-a-vis the previous model's line:

```{r}
y_binary = as.numeric(Xy_simple$response == 1)

svm_model_weights = linear_svm_learning_algorithm(cbind(Xy_simple$first_feature, Xy_simple$second_feature), y_binary)
my_svm_line = geom_abline(
    intercept = -svm_model_weights[1] / svm_model_weights[3],#NOTE: negative sign removed from intercept argument here
    slope = -svm_model_weights[2] / svm_model_weights[3], 
    color = "brown")
simple_viz_obj  + my_svm_line
```





Is this the same as what the `e1071` implementation returned? Why or why not?

TO-DO

## Multinomial Classification using KNN

Write a k = 1 nearest neighbor algorithm using the Euclidean distance function. The following comments are standard "Roxygen" format for documentation. Hopefully, we will get to packages at some point and we will go over this again. It is your job also to fill in this documentation.

```{r}
#' TO-DO: Provide a name for this function 
#' 1 nearest neighbor classification using Eculidean Distance
#' 
#' TO-DO: Explain what this function does in a few sentences
#' This function performs 1-Nearest Neighbor classification using the Euclidean distance metric. 
#' For each test observation in Xtest, it finds the closest observation in the training data Xinput
#' and assigns the corresponding class label from y_binary.
#' 
#'
#' @param Xinput      TO-DO: explain this argument  - The training data features as an n x p matrix.
#' @param y_binary    TO-DO: explain this argument  - The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param Xtest       TO-DO: explain this argument  - The testing data features as n x p matrix
#' @return            TO-DO: explain this argument  - the testing data responses as a vector of length n consisting of only 0's and 1's.
nn_algorithm_predict = function(Xinput, y_binary, Xtest){
  #TO-DO
  n_train = nrow(Xinput) 
  n_test = nrow(Xtest)
  y_test = array(NA, dim = n_test)
  
  for (i_test in 1:n_test){
    x_star = Xtest[i_test,]
    # 1 for row 2 for column in 2nd parameter of apply function
    distances = apply(Xinput, 1, function(x) {sum(x - x_star)^2})
    nearest_index = which.min(distances)
    y_test[i_test] = y_binary[nearest_index]
  }
    
    
    
  y_test
}
```

Write a few tests to ensure it actually works:

```{r}
#TO-DO
# Create a small training set with two features
X_train1 = matrix(c(0, 0,
                    1, 1,
                    2, 2), ncol = 2, byrow = TRUE)

y_train1 = c("A", "B", "B")


X_test1 = matrix(c(0.1, -0.1,  # close to (0,0)
                   1.2, 1.1,   # close to (1,1)
                   1.8, 2.1),  # close to (2,2)
                 ncol = 2, byrow = TRUE)


predictions = nn_algorithm_predict(X_train1, y_train1, X_test1)
print(predictions)


```






We now add an argument `d` representing any legal distance function to the `nn_algorithm_predict` function. Update the implementation so it performs NN(nearest neighbour) using that distance function. Set the default function to be the Euclidean distance in the original function. Also, alter the documentation in the appropriate places.

```{r}
#TO-DO
nn_algorithm_predict = function(Xinput, y_binary, Xtest, d = NULL){

  #TO-DO
  if (is.null(d)) {
    d = function(u, v) {sum((u-v)^2)}
  }
  
  n_train = nrow(Xinput) 
  n_test = nrow(Xtest)
  y_test = array(NA, dim = ntest)
  
  for (i_test in 1:n_test){
    x_star = Xtest[i_test,]
    # 1 for row 2 for column in 2nd parameter of apply function
    distances = apply(Xinput, 1, function(x) {d(x, x_star)})
    nearest_index = which.min(distances)
    y_test[i_test] = y_binary[nearest_index]
  }
    
    
    
  y_test
}
```

For extra credit (unless you're a masters student), add an argument `k` to the `nn_algorithm_predict` function and update the implementation so it performs KNN. In the case of a tie, choose yhat randomly. Set the default `k` to be the square root of the size of D which is an empirical rule-of-thumb popularized by the "Pattern Classification" book by Duda, Hart and Stork (2007). Also, alter the documentation in the appropriate places.

```{r}
#TO-DO for the 650 students but extra credit for undergrads
```

## Regression via OLS with one feature

Let's quickly recreate the sample data set from practice lecture 7:

```{r}
set.seed(1984)
n = 20
x = runif(n)
beta_0 = 3
beta_1 = -2
```

Compute h^* as `h_star_x`, then draw epsilon from an iid N(0, 0.33^2) distribution as `epsilon`, then compute the vector y.

```{r}
h_star_x = beta_0 + beta_1 * x
epsilon = rnorm(n, mean = 0, sd = 0.33)
y = h_star_x + epsilon
```

Graph the data by running the following chunk:

```{r}
pacman::p_load(ggplot2)
simple_df = data.frame(x = x, y = y)
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
simple_viz_obj
```

Does this make sense given the values of beta_0 and beta_1?
Yes. this make sense. as the model is h*(x) = 3 - 2x with the epsilon with the standard deviation of 0.33. the dots are going down as the slop is negative.

Write a function `my_simple_ols` that takes in a vector `x` and vector `y` and returns a list that contains the `b_0` (intercept), `b_1` (slope), `yhat` (the predictions), `e` (the residuals), `SSE`, `SST`, `MSE`, `RMSE` and `Rsq` (for the R-squared metric). Internally, you can only use the functions `sum` and `length` and other basic arithmetic operations. You should throw errors if the inputs are non-numeric or not the same length. You should also name the class of the return value `my_simple_ols_obj` by using the `class` function as a setter. No need to create ROxygen documentation here.

```{r}
my_simple_ols = function(x, y){
  ols_obj = list()
  #TO-DO
  if (!is.numeric(x) | !is.numeric(y)) {
    stop("Inputs must be numeric.")
  }
  if (length(x) != length(y)) {
    stop("Inputs must be the same length.")
  }
  
  n = length(x)
  
  x_bar = sum(x) / n
  y_bar = sum(y) / n
  
  numerator = sum(x*y) - (n * x_bar * y_bar)
  denomenator = sum(x^2) - (n * (x_bar) ^ 2)
  b_1 = numerator / denomenator
  
  b_0 = y_bar - b_1 * x_bar
  
  y_hat = b_0 + b_1 * x
  e = y - y_hat
  
  SSE = sum(e^2)
  SST = sum((y - y_bar)^2)
  
  MSE = SSE / (n-2)
  RMSE = sqrt(MSE)
  
  Rsq = 1 - SSE / SST
  
  ols_obj = list(b_0 = b_0,
                 b_1 = b_1,
                 e = e,
                 SSE = SSE,
                 MSE = MSE,
                 RMSE = RMSE,
                 Rsq = Rsq)
  
  class(ols_obj) = "my_simple_ols_obj"
  ols_obj
}
```

Verify your computations are correct for the vectors `x` and `y` from the first chunk using the `lm` function in R:

```{r}
lm_mod = lm(y ~ x)
my_simple_ols_mod = my_simple_ols(x, y)
#run the tests to ensure the function is up to spec
pacman::p_load(testthat)
expect_equal(my_simple_ols_mod$b_0, as.numeric(coef(lm_mod)[1]), tol = 1e-4)
expect_equal(my_simple_ols_mod$b_1, as.numeric(coef(lm_mod)[2]), tol = 1e-4)
expect_equal(my_simple_ols_mod$RMSE, summary(lm_mod)$sigma, tol = 1e-4)
expect_equal(my_simple_ols_mod$Rsq, summary(lm_mod)$r.squared, tol = 1e-4)
```

Verify that the average of the residuals is 0 using the `expect_equal`. Hint: use the syntax above.

```{r}
#TO-DO

expect_equal(sum(my_simple_ols_mod$e) / length(my_simple_ols_mod$e), 0, tol = 1e-4)

```

Create the $X$ matrix for this data example. Make sure it has the correct dimension.

```{r}
#TO-DO
X = cbind(1,x)
X
```

Use the `model.matrix` function to compute the matrix `X` and verify it is the same as your manual construction.

```{r}
#TO-DO
X = model.matrix(~x)
X
```

Create a prediction method `g` that takes in a vector `x_star` and `my_simple_ols_obj`, an object of type `my_simple_ols_obj` and predicts y values for each entry in `x_star`. 

```{r}
g = function(my_simple_ols_obj, x_star){
  #TO-DO
  y_hat_star = my_simple_ols_obj$b_0 + my_simple_ols_obj$b_1 * x_star
  return (y_hat_star)
}
```

Use this function to verify that when predicting for the average x, you get the average y.

```{r}
expect_equal(g(my_simple_ols_mod, mean(x)), mean(y))
```


In class we spoke about error due to ignorance, misspecification error and estimation error. Show that as n grows, estimation error shrinks. Let us define an error metric that is the difference between b_0 and b_1 and beta_0 and beta_1. How about ||b - beta||^2 where the quantities are now the vectors of size two. Show as n increases, this shrinks. 

```{r}
beta_0 = 3
beta_1 = -2
beta = c(beta_0, beta_1)

ns = c(20, 50, 100, 200, 500, 1000, 5000, 10000)
errors = array(NA, length(ns))
for (i in 1 : length(ns)) {
  n = ns[i]
  x = runif(n)
  h_star_x = beta_0 + beta_1 * x 
  epsilon = rnorm(n, mean = 0, sd = 0.33)
  y = h_star_x + epsilon
  
  #TO-DO
  mod = my_simple_ols(x,y)
  b = c(mod$b_0, mod$b_1)
  errors[i] = sum((beta - b)^2)
}
errors
```


We are now going to repeat one of the first linear model building exercises in history --- that of Sir Francis Galton in 1886. First load up package `HistData`.

```{r}
pacman::p_load(HistData)
```

In it, there is a dataset called `Galton`. Load it up.

```{r}
data(Galton)
```

You now should have a data frame in your workspace called `Galton`. Summarize this data frame and write a few sentences about what you see. Make sure you report n, p and a bit about what the columns represent and how the data was measured. See the help file `?Galton`.
p is 1 and n is 928 the number of observations  

```{r}
pacman::p_load(skimr)
skim(Galton)
?Galton
```

TO-DO
The data frame has n = 928 and p = 1. There are 928 rows and 1 columns. The columns represent the height of the children and an average height of father and mother. The data were taken in the 19th century, and Galton was interested in the hereditary patterns of stature.

Find the average height (include both parents and children in this computation).

```{r}
#TO-DO
y_bar = mean(Galton$child)
x_bar = mean(Galton$parent)
y_bar
x_bar
```

If you were predicting child height from parent and you were using the null model, what would the RMSE be of this model be?

```{r}
#TO-DO
null_RMSE = sqrt(sum((Galton$child - mean(Galton$child))^2) / nrow(Galton))
null_RMSE
```

Note that in Math 241 you learned that the sample average is an estimate of the "mean", the population expected value of height. We will call the average the "mean" going forward since it is probably correct to the nearest tenth of an inch with this amount of data.

Run a linear model attempting to explain the childrens' height using the parents' height. Use `lm` and use the R formula notation. Compute and report b_0, b_1, RMSE. 

```{r}
#TO-DO
mod = lm(child ~ parent, data = Galton)
b_0 = mod$coefficients[1]
b_1 = mod$coefficients[2]
RMSE = summary(mod)$sigma

b_0
b_1
RMSE
```

Interpret all four quantities: b_0, b_1, RMSE and R^2. Use the correct units of these metrics in your answer.

TO-DO
- b_0 is the intercept and b_1 is the slope. The units are in inches so if the parents increase in one inch, the children height increase in about 0.65 inch. RMSE is the average deviation of the child heights from the predicted by the model which means there are about 2 inches varation in prediction.

How good is this model? How well does it predict? Discuss.

TO-DO
- This model is not bad and a decent one. If the parent height increases, the children height will increase too. The error 2 inches is a reasonable variance among heights.


It is reasonable to assume that parents and their children have the same height? Explain why this is reasonable using basic biology and common sense.

TO-DO
- Yes. It is reasonable to assume that parents and their children have the same height. In common sense and a gene, if the parents have high height, their children will have high height and vise visa. They could have same height as they have same genetics and shared environmental influences. 

If they were to have the same height and any differences were just random noise with expectation 0, what would the values of beta_0 and beta_1 be?

TO-DO
beta_0 would be 0 and beta_1 would be 1. 

Let's plot (a) the data in D as black dots, (b) your least squares line defined by b_0 and b_1 in blue, (c) the theoretical line beta_0 and beta_1 if the parent-child height equality held in red and (d) the mean height in green.

```{r}
pacman::p_load(ggplot2)
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b_0, slope = b_1, color = "blue", size = 1) +
  geom_abline(intercept = 0, slope = 1, color = "red", size = 1) +
  geom_abline(intercept = y_bar, slope = 0, color = "darkgreen", size = 1) +
  xlim(63.5, 72.5) + 
  ylim(63.5, 72.5) +
  coord_equal(ratio = 1)
```

Fill in the following sentence: 

TO-DO: Children of short parents became taller on average and children of tall parents became shorter on average.

Why did Galton call it "Regression towards mediocrity in hereditary stature" which was later shortened to "regression to the mean"?

TO-DO
- because the heights with the extremely short and tall parents are close to the average height which mean tall parents tended to have somewhat shorter children and short parents tended to have somewhat taller children. Children's heights were not as extreme as their parents' heights.


Why should this effect be real?


TO-DO
- this effect should be real because the height is not always one to one ratio. Even the parents have extreme tall or short, the average of the children is not extreme. In term of genetics, the height of the parent and the children will not always be mirror.


You now have unlocked the mystery. Why is it that when modeling with y continuous, everyone calls it "regression"? Write a better, more descriptive and appropriate name for building predictive models with y continuous.

TO-DO
- It calls regression because it regresses toward to average when it becomes extreme values. I would say Continuity Model.

You can now clear the workspace.

```{r}
rm(list = ls())
```

Create a dataset D which we call `Xy` such that the linear model has R^2 about 50\% and RMSE approximately 1.

```{r}
x = #TO-DO
y = #TO-DO
Xy = data.frame(x = x, y = y)
```

Extra credit but required for 650 students: create a dataset D and a model that can give you R^2 arbitrarily close to 1 i.e. approximately 1 - epsilon but RMSE arbitrarily high i.e. approximately M.

```{r}
epsilon = 0.01
M = 1000
#TO-DO
```






