---
title: "Lab 10"
author: "Ye Htut Maung"
output: pdf_document
---

#YARF

For the next couple of labs, I want you to make some use of a package I wrote that offers convenient and flexible tree-building and random forest-building. Make sure you have RTools installed (if you're on windows) and then a JDK installed:

https://www.oracle.com/java/technologies/downloads/

Then try to install rJava

```{r}
options(java.parameters = "-Xmx8000m")
pacman::p_load(rJava)
.jinit()
```

```{r}
remove.packages("YARF")
pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
options(java.parameters = "-Xmx8000m")
pacman::p_load(YARF)
```

If you have error, messages, try to google them. Everyone has trouble with rJava!

If that worked, please try to run the following which will install YARF from my github:

```{r}
if (!pacman::p_isinstalled(YARF)){
  pacman::p_install_gh("kapelner/YARF/YARFJARs", ref = "dev")
  pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
}
pacman::p_load(YARF)
```

Please try to fix the error messages (if they exist) as best as you can. I can help on slack.

# Regression Trees

You can use the `YARF` package if it works, otherwise, use the `randomForest` package (the canonical R package for this algorithm).

Let's take a look at a simulated sine curve. Below is the code for the data generating process:

```{r}
rm(list = ls())
n = 500
sigma = 0.3
x_min = 0
x_max = 10

f_x = function(x){sin(x)}
x = runif(n, x_min, x_max)
y = f_x(x) + rnorm(n, 0, sigma)

options(java.parameters = "-Xmx4000m")
pacman::p_load(YARF, tidyverse, magrittr)
```

Provide the bias-variance decomposition of this DGP fit with the tree model fitted with the optimal N_0 from the previous lab. It is a lot of code, but it is in the practice lectures. If your three numbers don't add up within two significant digits, increase your resolution.

```{r}
#TO-DO
library(YARF)

# Set parameters
n_train = 500
n_test = 500
sigma = 0.3
Nsim = 1000
x_min = 0
x_max = 10
f_x = function(x){sin(x)}

# Storage for models and residuals
training_gs = list()
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

# Loop over simulations
for (nsim in 1:Nsim){
  # Training data
  x_train = runif(n_train, x_min, x_max)
  y_train = f_x(x_train) + rnorm(n_train, 0, sigma)
  
  # Fit tree model with optimal N0 = 5
  g_model = YARFCART(data.frame(x = x_train), y_train, 
                     calculate_oob_error = FALSE, 
                     verbose = FALSE, nodesize = 5)  
  training_gs[[nsim]] = g_model
  
  # Test data
  x_test = runif(n_test, x_min, x_max)
  y_test = f_x(x_test) + rnorm(n_test, 0, sigma)
  
  y_hat = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat
}


ggplot(data.frame(x = x_train, y = y_train)) + 
  geom_point(aes(x, y))

head(training_gs, 2)

mse = mean(all_residuals^2)
mse

resolution = 1000
x_grid = seq(x_min, x_max, length.out = resolution)
g_avg = rep(0, resolution)

for (nsim in 1:Nsim){
  g_nsim = training_gs[[nsim]]
  g_avg = g_avg + predict(g_nsim, data.frame(x = x_grid))
}
g_avg = g_avg / Nsim
biases = f_x(x_grid) - g_avg
expe_bias_g_sq = mean(biases^2)

ggplot(data.frame(x = x_grid, f = f_x(x_grid), expe_g = g_avg)) + 
  geom_line(aes(x, expe_g), col = "red", lwd = 2) +
  geom_line(aes(x, f), col = "green", lwd = 1)



# Plot up to 30 individual tree fits + average + true function
plot_obj = ggplot(data.frame(x = x_grid)) + 
  xlim(x_min, x_max)

num_trees_to_visualize = 30

for (nsim in 1:min(Nsim, num_trees_to_visualize)) {
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x_grid))
  plot_obj = plot_obj + 
    geom_line(data = data.frame(x = x_grid, y = g_x), 
              aes(x = x, y = y), 
              col = "blue", alpha = 0.3)
}

# Add the expected prediction (red) and true function (green)
plot_obj + 
  geom_line(data = data.frame(x = x_grid, expe_g = g_avg), 
            aes(x = x, y = expe_g), 
            col = "red", lwd = 2) +
  geom_line(data = data.frame(x = x_grid, f = f_x(x_grid)), 
            aes(x = x, y = f), 
            col = "green", lwd = 1) 

var_x_s = array(NA, min(Nsim, 50))
for (nsim in 1 : min(Nsim, 50)){
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x_grid))
  var_x_s[nsim] = mean((g_x - g_avg)^2)
}
expe_var_g = mean(var_x_s)
expe_var_g


mse
sigma^2
expe_bias_g_sq
expe_var_g
sigma^2 + expe_bias_g_sq + expe_var_g
```

# Classification Trees

Let's get the letter recognition data from the `mlbench` package.

```{r}
rm(list = ls())
pacman::p_load(mlbench)
data(LetterRecognition, package = "mlbench")
n = nrow(LetterRecognition)
skimr::skim(LetterRecognition)
```

This dataset has 20,000 examples. Create a training-select-test split so that they each have 1,000 observations.

```{r}
train_idx = sample(1 : n, 1000)
select_idx = sample(setdiff(1 : n, train_idx), 1000)
test_idx = sample(setdiff(1 : n, c(train_idx, select_idx)), 1000)
letters_train = LetterRecognition[train_idx, ]
letters_select = LetterRecognition[select_idx, ]
letters_test = LetterRecognition[test_idx, ]

```

Find the optimal classification tree by using the model selection algorithm to optimize the nodesize hyperparameter. Use misclassification error as the performance metric.

```{r}
nodesizes = seq(1, 20, by = 1)
misclassification_errs = array(NA, length(nodesizes))
for (m in 1:length(nodesizes)){
  tree_mod = YARFCART(letters_train[, 2:17], letters_train[, 1], calculate_oob_error = FALSE, nodesize = nodesizes[m])
  misclassification_errs[m] = sum(predict(tree_mod, letters_select[, 2:17]) != letters_select[, 1])
}
```

Plot the oos misclassification error by nodesize.

```{r}
ggplot(data.frame(nodesize = nodesizes, misclassification_error = misclassification_errs)) + 
  aes(x = nodesize, y = misclassification_error) +
  geom_point() + 
  geom_line()
```

Construct the optimal classification tree on train and select sets. Then estimate generalization error. Save `y_hat_test` as we'll need it later.

```{r}
tree_mod_opt = YARFCART(rbind(letters_train, letters_select)[, 2:17], rbind(letters_train, letters_select)[, 1], calculate_oob_error = FALSE, nodesize = nodesizes[which.min(misclassification_errs)])


```

Print out the top of the tree so we can have some level of interpretation to how the model g is predicting.

```{r}
illustrate_trees(tree_mod_opt, max_depth = 5, length_in_px_per_half_split = 30, open_file = TRUE)
```

Create a "confusion matrix". This means it shows every predicted level (which is a letter in our case) and every actual level. Here you'll see every type of error e.g. "P was predicted but the real letter is H", "M was predicted but the real letter is N" etc. This is really easy: one call to the `table` function is all you need.

```{r}
#TO-DO
y_oos = predict(tree_mod_opt, letters_test[, 2: 17])
table(letters_test[, 1], y_oos)
```

Which errors are most prominent in this model?

#TO-DO
P and F, Q and O, V and Y, W and N, W and Q are most prominent in this model because thoses letters had the most mismatches.

#Bagged Trees and Random Forest

Take a training sample of n = 2000 observations from the diamonds data.

```{r}
rm(list = ls())
pacman::p_load(tidyverse)
set.seed(1)
diamonds_train = ggplot2::diamonds %>% 
  sample_n(2000)
```


Using the diamonds data, find the oob s_e for a bagged-tree model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. If you are using the `randomForest` package, you can create the bagged tree model via setting an argument within the RF constructor function. Plot.

```{r}
num_trees_values = c(1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000)
oob_se_bagged_trees_mod_by_num_trees = array(NA, length(num_trees_values))
for (m in 1 : length(num_trees_values)) {
  bag_mod = YARFBAG(diamonds_train[, -7], diamonds_train$price, num_trees = num_trees_values[m])
  oob_se_bagged_trees_mod_by_num_trees[m] = bag_mod$rmse_oob
}
pacman::p_load(ggplot2)
ggplot(data.frame(num_trees_values = num_trees_values, oob_err = oob_se_bagged_trees_mod_by_num_trees)) +
  geom_point(aes(x = num_trees_values, y = oob_err))+
  geom_line(aes(x = num_trees_values, y = oob_err))+
  scale_x_log10()+
  scale_y_log10()
  
```

Find the bootstrap s_e for a RF model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. If you are using the `randomForest` package, you can calculate oob residuals via `e_oob = y_train - rf_mod$predicted`. Plot.

```{r}
oob_se_rf_mod_by_num_trees = array(NA, length(num_trees_values))
for (m in 1 : length(num_trees_values)) {
  bag_mod = YARF(diamonds_train[, -7], diamonds_train$price, num_trees = num_trees_values[m])
  oob_se_rf_mod_by_num_trees[m] = bag_mod$rmse_oob
}
ggplot(data.frame(num_trees_values = num_trees_values, oob_err = oob_se_rf_mod_by_num_trees)) +
  geom_point(aes(x = num_trees_values, y = oob_err))+
  geom_line(aes(x = num_trees_values, y = oob_err))+
  scale_x_log10()+
  scale_y_log10()
```

What is the percentage gain / loss in performance of the RF model vs bagged trees model for each number of trees? Gains are negative (as in lower oos s_e).

```{r}
df = data.frame(
  num_trees = num_trees_values,
  rf_improvement = (oob_se_rf_mod_by_num_trees - oob_se_bagged_trees_mod_by_num_trees) / oob_se_bagged_trees_mod_by_num_trees * 100
)
df
```

Why was this the result?

#TODO
Bagged model is better than Random forest model because most features helps reduce bias without much overfitting and randomness in Random Forests reduced accuracy by limiting useful feature splits.

Plot oob s_e by number of trees for both RF and bagged trees by creating a long data frame from the two results.

```{r}
#TO-DO


df_long = data.frame(
  num_trees = num_trees_values,
  Bagged = oob_se_bagged_trees_mod_by_num_trees,
  RandomForest = oob_se_rf_mod_by_num_trees
)


df_long = df_long %>%
  pivot_longer(cols = c(Bagged, RandomForest), 
               names_to = "model_type", 
               values_to = "oob_se")


ggplot(df_long, aes(x = num_trees, y = oob_se, color = model_type)) +
  geom_point() +
  geom_line() +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "OOB Standard Error vs Number of Trees",
       x = "Number of Trees (log scale)", 
       y = "OOB Standard Error (log scale)", 
       color = "Model Type") +
  theme_minimal()
```

Build RF models for 500 trees using different `mtry` values: 1, 2, ... the maximum. That maximum will be the number of features assuming that we do not binarize categorical features if you are using `randomForest` or the number of features assuming binarization of the categorical features if you are using `YARF`. Calculate oob s_e for all mtry values.

```{r}
oob_se_by_mtry = array(NA, ncol(diamonds_train))
#TO-DO
p = ncol(diamonds_train[, -7])  # number of predictors after binarization
oob_se_by_mtry = array(NA, p)

# Loop through all mtry values from 1 to p
for (m in 1:p) {
  rf_mod = YARF(diamonds_train[, -7], diamonds_train$price, 
                num_trees = 500, mtry = m)
  oob_se_by_mtry[m] = rf_mod$rmse_oob
}


oob_se_by_mtry
```

Plot oob s_e by mtry.

```{r}
#TO-DO
df_mtry = data.frame(
  mtry = 1:length(oob_se_by_mtry),
  oob_se = oob_se_by_mtry
)


library(ggplot2)

ggplot(df_mtry, aes(x = mtry, y = oob_se)) +
  geom_point() +
  geom_line() +
  labs(
    title = "OOB Standard Error vs mtry",
    x = "mtry (Number of Features Tried at Each Split)",
    y = "OOB Standard Error"
  ) +
  theme_minimal()
```

Take a sample of n = 2000 observations from the adult data and name it `adult_sample`.  Then impute missing values using missForest (we will cover what this is later in.

```{r}
rm(list = ls())
pacman::p_load(missForest)
set.seed(1)
pacman::p_load_gh("coatless/ucidata")
adult_train = adult %>% 
  sample_n(2000)
adult_train[] = lapply(adult_train, function(x) {
  if (is.character(x)) factor(x) else x
})

adult_train = missForest(adult_train)$ximp
adult_train
```


Using the adult_train data, find the bootstrap misclassification error for a bagged-tree model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. Plot.

```{r}
num_trees_values = c(1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000)
oob_se_bagged_trees_mod_by_num_trees = array(NA, length(num_trees_values))
#TO-DO
pacman::p_load(randomForest)
adult_train$income = as.factor(adult_train$income)


for (i in seq_along(num_trees_values)) {
  model = randomForest(income ~ ., 
                       data = adult_train, 
                       ntree = num_trees_values[i],
                       mtry = ncol(adult_train) - 1)  
  oob_se_bagged_trees_mod_by_num_trees[i] = model$err.rate[num_trees_values[i], "OOB"]
}


df_plot = data.frame(
  num_trees = num_trees_values,
  oob_error = oob_se_bagged_trees_mod_by_num_trees
)

ggplot(df_plot, aes(x = num_trees, y = oob_error)) +
  geom_point() +
  geom_line() +
  labs(
    title = "OOB Misclassification Error vs Number of Trees (Bagged Trees)",
    x = "Number of Trees",
    y = "OOB Misclassification Error"
  ) +
  theme_minimal()
```

Using the adult_train data, find the bootstrap misclassification error for an RF model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees.

```{r}
oob_se_rf_mod_by_num_trees = array(NA, length(num_trees_values))
#TO-DO
p = ncol(adult_train) - 1  # subtract 1 for the response column
mtry_rf = floor(sqrt(p))


for (i in seq_along(num_trees_values)) {
  model = randomForest(income ~ ., 
                       data = adult_train,
                       ntree = num_trees_values[i],
                       mtry = mtry_rf)  # Random Forest uses a subset of predictors
  
  oob_se_rf_mod_by_num_trees[i] = model$err.rate[num_trees_values[i], "OOB"]
}

df_rf = data.frame(
  num_trees = num_trees_values,
  oob_error = oob_se_rf_mod_by_num_trees
)

ggplot(df_rf, aes(x = num_trees, y = oob_error)) +
  geom_point() +
  geom_line() +
  labs(
    title = "OOB Misclassification Error vs Number of Trees (Random Forest)",
    x = "Number of Trees",
    y = "OOB Misclassification Error"
  ) +
  theme_minimal()
```

What is the percentage gain / loss in performance of the RF model vs bagged trees model?

```{r}
cbind(
  num_trees_values,
  (oob_se_rf_mod_by_num_trees - oob_se_bagged_trees_mod_by_num_trees) / oob_se_bagged_trees_mod_by_num_trees * 100
)
```

Build RF models on adult_train for 500 trees using different `mtry` values: 1, 2, ... the maximum (see above as maximum is defined by the specific RF algorithm implementation). 

```{r}
oob_se_by_mtry = array(NA, ncol(adult_train)-1)
#TO-DO


p = ncol(adult_train) - 1 # subtract 1 for the response column


# Loop through mtry values from 1 to p
for (m in 1:p) {
  model = randomForest(income ~ ., 
                       data = adult_train, 
                       ntree = 500,
                       mtry = m)
  oob_se_by_mtry[m] = model$err.rate[500, "OOB"]
}

oob_se_by_mtry
```


Plot bootstrap misclassification error by `mtry`.

```{r}
#TO-DO
df_mtry = data.frame(
  mtry = 1:length(oob_se_by_mtry),
  oob_error = oob_se_by_mtry
)

# Plot
ggplot(df_mtry, aes(x = mtry, y = oob_error)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Bootstrap Misclassification Error vs mtry (500 Trees)",
    x = "mtry (Number of Features Tried at Each Split)",
    y = "OOB Misclassification Error"
  ) +
  theme_minimal()
```

Is `mtry` an important hyperparameter to optimize when using the RF algorithm? Explain

#TO-DO
Yes, mtry is important to optimize in Random Forests because it affects model accuracy. As shown in the plot, different mtry values lead to different OOB errors, with some values clearly performing better—highlighting its impact on generalization.

Identify the best model among all values of `mtry`. Fit this RF model. Then report the following oob error metrics: misclassification error, precision, recall, F1, FDR, FOR and compute a confusion matrix.

```{r}
#TO-DO
# Step 1: Identify best mtry and fit RF model
best_mtry = which.min(oob_se_by_mtry)
best_rf = randomForest(income ~ ., data = adult_train, ntree = 500, mtry = best_mtry)

# Step 2: Extract OOB predictions and true labels
y_test = adult_train$income
p_hats_oob = best_rf$votes[, ">50K"]
y_hats_oob = factor(ifelse(p_hats_oob >= 0.5, ">50K", "<=50K"), levels = levels(y_test))  # default 0.5 threshold

# Step 3: Misclassification error and confusion matrix
mean(y_hats_oob != y_test)
oob_conf_table = table(y_test, y_hats_oob)
oob_conf_table

n = sum(oob_conf_table)
fp = oob_conf_table[1, 2]
fn = oob_conf_table[2, 1]
tp = oob_conf_table[2, 2]
tn = oob_conf_table[1, 1]

num_pred_pos = sum(oob_conf_table[, 2])
num_pred_neg = sum(oob_conf_table[, 1])
num_pos = sum(oob_conf_table[2, ])
num_neg = sum(oob_conf_table[1, ])

# Precision
precision = tp / num_pred_pos
cat("precision", round(precision * 100, 2), "%\n")

# Recall
recall = tp / num_pos
cat("recall", round(recall * 100, 2), "%\n")

# F1 Score
f1 = 2 * precision * recall / (precision + recall)
cat("F1 score", round(f1 * 100, 2), "%\n")

# False Discovery Rate
false_discovery_rate = 1 - precision
cat("FDR", round(false_discovery_rate * 100, 2), "%\n")

# False Omission Rate
false_omission_rate = fn / num_pred_neg
cat("FOR", round(false_omission_rate * 100, 2), "%\n")

# Misclassification Error
misclass_error = mean(y_hats_oob != y_test)
cat("Misclassification Error", round(misclass_error * 100, 2), "%\n")
```

Is this a good model? (yes/no and explain).

#TO-DO
I would say it is not a bad model but not too good. Because we have low misclassification error (15.3%), high precision (73.66%), low FDR(26.34% ) and low FOR (12.45%) but recall is moderate(60.4%) and F1 score is moderate (66.37%).

There are probability asymmetric costs to the two types of errors. Assign two costs below and calculate oob total cost.

```{r}
fp_cost = 10
fn_cost = 50
#TO-DO
# Use previously computed confusion matrix: oob_conf_table
fp = oob_conf_table[1, 2]
fn = oob_conf_table[2, 1]

# Calculate total cost
total_cost = (fp_cost * fp) + (fn_cost * fn)
cat("OOB Total Cost:", total_cost, "\n")
```

# Missing Data

Load up the Boston Housing Data and separate into matrix `X` for the features and vector `y` for the response. Randomize the rows

```{r}
rm(list = ls())
set.seed(1)
boston = MASS::Boston
boston_shuffled = boston[sample(1 : nrow(boston)), ]
X = as.matrix(boston_shuffled[, 1 : 13])
y = boston_shuffled$medv
rm(boston, boston_shuffled)
```



Similar to lab 1, write a function that takes a matrix and punches holes (i.e. sets entries equal to `NA`) randomly with an argument `prob_missing`.

```{r}
punch_holes = function(mat, prob_missing){
  n = nrow(mat) * ncol(mat)
  is_missing = as.logical(rbinom(n, 1, prob_missing))
  mat[is_missing] = NA
  mat
}
```

Create a matrix `Xmiss` which is `X` but has missingness with probability of 10% using the function you just wrote. 

```{r}
Xmiss = punch_holes(X, 0.1)
```

Also, generate the M matrix and delete columns that have no missingness.

```{r}
M = apply(is.na(Xmiss), 2, as.numeric)
colnames(M) = paste("is_missing_", colnames(X), sep = "")
M = M[, colSums(M) > 0]
```

Split the first 400 observations were the training data and the remaining observations are the test set. For Xmiss, cbind on the M so the model has a chance to fit on "is missing" as we discussed in class.

```{r}
train_idx = 1 : 400
test_idx = setdiff(1 : nrow(X), train_idx)
X_train =     X[train_idx, ]
Xmiss_train = cbind(Xmiss, M)[train_idx, ]
y_train =     y[train_idx]
X_test =      X[test_idx, ]
Xmiss_test =  cbind(Xmiss, M)[test_idx, ]
y_test =      y[test_idx]
```

Fit a random forest model of `y_train ~ X_train`, report oos s_e (not oob) on `X_test`. This ignores missingness

```{r}
rf_mod = YARF(data.frame(X_train), y_train, use_missing_data = TRUE)
y_hat_test = predict(rf_mod, data.frame(X_test))
rmse_ignore_missingness = sqrt(mean((y_hat_test - y_test)^2))
rmse_ignore_missingness
```

Impute the missingness in `Xmiss` using the feature averages to create a matrix `Ximp_naive_train` and `Ximp_naive_test`. 

```{r}
#TO-DO
col_means = colMeans(Xmiss_train, na.rm = TRUE)

# Impute training set
Ximp_naive_train = Xmiss_train
for (j in seq_along(col_means)) {
  na_idx = is.na(Ximp_naive_train[, j])
  Ximp_naive_train[na_idx, j] = col_means[j]
}

# Impute test set using same training means
Ximp_naive_test = Xmiss_test
for (j in seq_along(col_means)) {
  na_idx = is.na(Ximp_naive_test[, j])
  Ximp_naive_test[na_idx, j] = col_means[j]
}

Ximp_naive_test
Ximp_naive_train
```

Fit a random forest model of `y_train ~ Ximp_naive_train`, report oos s_e (not oob) on `Ximp_naive_test`.

```{r}
#TO-DO
# Fit RF model using imputed training data
rf_mod_naive = YARF(data.frame(Ximp_naive_train), y_train)

# Predict on imputed test data
y_hat_test = predict(rf_mod_naive, data.frame(Ximp_naive_test))
rmse_naive_impute = sqrt(mean((y_hat_test - y_test)^2))
rmse_naive_impute
```

How much predictive performance was lost due to missingness when naive imputation was used vs when there was no missingness?

```{r}
#TO-DO

100 * (rmse_naive_impute - rmse_ignore_missingness) / rmse_ignore_missingness
#about 20% lose
```

Use `missForest` to impute the missing entries to create a matrix `Ximp_MF_train` and `Ximp_MF_test`.

```{r}
pacman::p_load(missForest)
Xy_miss = rbind(
  cbind(Xmiss_train, y_train),
  cbind(Xmiss_test, NA)
)

Xy_missimp = missForest(Xy_miss)$ximp
Ximp_MF_train = Xy_missimp[1:nrow(X_train), 1:(ncol(Xy_missimp)-1)]
Ximp_MF_test = Xy_missimp[(nrow(X_train)+1):nrow(Xy_missimp), 1:(ncol(Xy_missimp)-1)]
```

Fit a random forest model of `y_train ~ Ximp_MF_train`, report oos s_e (not oob) on `Ximp_MF_test`.

```{r}
#TO-DO
rf_mod = YARF(data.frame(Ximp_MF_train), y_train)
y_hat_test = predict(rf_mod, data.frame(Ximp_MF_test))
rmse_mf_impute = sqrt(mean((y_hat_test - y_test)^2))
rmse_mf_impute
```

How much predictive performance was lost due to missingness when `missForest` imputation was used?

```{r}
#TO-DO


100 * (rmse_mf_impute - rmse_ignore_missingness) / rmse_ignore_missingness

# only about 1% or less loss
```

Why did `missForest` imputation perform better than naive imputation?

#TO-DO
missForest performed better than naive imputation because it uses information from all other features to intelligently predict missing values using Random Forests, while naive imputation simply fills missing values with column averages, ignoring relationships between variables.


Reload the feature matrix:

```{r}
rm(list = ls())
X = as.matrix(MASS::Boston[, 1 : 13])
```
