---
title: "Lab 11"
author: "Ye Htut Maung"
output: pdf_document
---


# Asymmetric Cost Modeling, ROC and DET curves

Load the adult dataset and impute the missing data using the `missForest` package.

```{r}
pacman::p_load_gh("coatless/ucidata")
data(adult)


pacman::p_load(missForest, dplyr)
#TO-DO
adult = na.omit(adult)
n = nrow(adult)
js = sapply(adult, is.character)
adult[, js] = lapply(adult[, js], as.factor)
adult = missForest(adult, sampsize = rep(1000, ncol(adult)))$ximp
adult$income = ifelse(adult$income == ">50K", 1, 0)
```

Then sample training and testing sets each of size 2,000:

```{r}

set.seed(1984)
n_subset = 2000

pacman::p_load(skimr)

all_represented = FALSE
while (!all_represented) {
  index_train = sample.int(n, n_subset, replace = FALSE)
  adult_train = adult[index_train,]
  
  all_represented = TRUE
  for (j in 1 : ncol(adult_train)) {
    col = adult_train[, j]
    if (!is.factor(col)) {
      next
    }
    
    if (n_unique(col) != length(levels(col))) {
      all_represented = FALSE
      break
    }
  }
}

index_test = setdiff(1 : n, index_train)
adult_test = adult[index_test,]


```


Fit a logistic regression model to the adult_train missingness-imputed data.

```{r}
#TO-DO
prob_mob = glm(income ~ ., data = adult_train, family = "binomial")
```

Use the function from class to calculate all the error metrics (misclassification error, precision, recall, F1, FDR, FOR) for the values of the probability threshold being 0.001, 0.002, ..., 0.999 in a tibble (dplyr data frame).



```{r}
pacman::p_load(tidyverse)
p_hats = predict(prob_mob, adult_train, type = "response")

compute_metrics_prob_classifier = function(p_hats, y_true, res = 0.001){
  #we first make the grid of all prob thresholds
  p_thresholds = seq(0 + res, 1 - res, by = res) #values of 0 or 1 are trivial
  
  #now we create a matrix which will house all of our results
  performance_metrics = matrix(NA, nrow = length(p_thresholds), ncol = 12)
  colnames(performance_metrics) = c(
    "p_th",
    "TN",
    "FP",
    "FN",
    "TP",
    "miscl_err",
    "precision",
    "recall",
    "FDR",
    "FPR",
    "FOR",
    "miss_rate"
  )
  
  #now we iterate through each p_th and calculate all metrics about the classifier and save
  n = length(y_true)
  for (i in 1 : length(p_thresholds)){
    p_th = p_thresholds[i]
    y_hats = factor(ifelse(p_hats >= p_th, 1, 0))
    confusion_table = table(
      factor(y_true, levels = c(0, 1)),
      factor(y_hats, levels = c(0, 1))
    )
      
    fp = confusion_table[1, 2]
    fn = confusion_table[2, 1]
    tp = confusion_table[2, 2]
    tn = confusion_table[1, 1]
    npp = sum(confusion_table[, 2])
    npn = sum(confusion_table[, 1])
    np = sum(confusion_table[2, ])
    nn = sum(confusion_table[1, ])
  
    performance_metrics[i, ] = c(
      p_th,
      tn,
      fp,
      fn,
      tp,
      (fp + fn) / n,
      tp / npp, #precision
      tp / np,  #recall
      fp / npp, #false discovery rate (FDR)
      fp / nn,  #false positive rate (FPR)
      fn / npn, #false omission rate (FOR)
      fn / np   #miss rate
    )
  }
  
  #finally return the matrix
  performance_metrics
}

asymmetric_predictions_results = as_tibble(compute_metrics_prob_classifier(p_hats, adult_train$income))
head(asymmetric_predictions_results)
#TO-DO

```

Calculate the column `total_cost` and append it to this data frame via `mutate`.

```{r}
#TO-DO
c_FP = 200
c_FN = 1000
asymmetric_predictions_results = asymmetric_predictions_results %>%
  mutate(total_cost = c_FP * FP + c_FN * FN)
```

Which is the lowest total cost? What is the "winning" probability threshold value providing that minimum total cost?

```{r}
#TO-DO
min(asymmetric_predictions_results$total_cost)
asymmetric_predictions_results$p_th[which.min(asymmetric_predictions_results$total_cost)]
```

Plot an ROC curve in-sample and interpret.

```{r}
#TO-DO
pacman::p_load(ggplot2)
ggplot(asymmetric_predictions_results) +
  geom_line(aes(x = FPR, y = recall, col = p_th), lwd = 2) +
  geom_abline(intercept = 0, slope = 1) + 
  coord_fixed() + xlim(0, 1) + ylim(0, 1) + 
  scale_colour_gradientn(colours = rainbow(5))
```

#TO-DO interpretation


This ROC curve shows the trade-off between recall and false positive rate as the classification threshold p_th changes (color-coded). The curve lies above the diagonal, meaning the model performs better than random. Lower thresholds (red) give high recall but high FPR; higher thresholds (blue) reduce FPR but also recall.
Calculate AUC in-sample and interpret.

```{r}
#TO-DO
pacman::p_load(pracma)
-trapz(asymmetric_predictions_results$FPR, asymmetric_predictions_results$recall)
```

#TO-DO interpretation
The AUC of 0.78 indicates that the model has good discriminatory ability. It means that, on average, the model ranks a randomly chosen positive instance higher than a randomly chosen negative instance 78% of the time.

Plot a FDR-FOR curve in-sample and interpret.

```{r}
#TO-DO
ggplot(asymmetric_predictions_results) +
  geom_line(aes(x = FDR, y = FOR)) +
  coord_fixed() + 
  xlim(0, max(asymmetric_predictions_results$FDR)) + 
  ylim(0, max(asymmetric_predictions_results$FOR))
```

Plot an ROC curve oos and interpret.

```{r}
#TO-DO


# Predicted probabilities on the test set
p_hats_test = predict(prob_mob, adult_test, type = "response")

# Compute performance metrics using the provided function
asymmetric_predictions_results_oos = as_tibble(
  compute_metrics_prob_classifier(p_hats_test, adult_test$income)
)

# Plot the ROC curve
ggplot(asymmetric_predictions_results_oos) +
  geom_line(aes(x = FPR, y = recall, color = p_th), linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_fixed() +
  xlim(0, 1) + ylim(0, 1) +
  scale_color_gradientn(colours = rainbow(5)) +
  labs(
    title = "ROC Curve (Out-of-Sample)",
    x = "False Positive Rate (FPR)",
    y = "True Positive Rate (Recall)",
    color = "Probability Threshold"
  )
```

#TO-DO interpretation
This out-of-sample ROC curve shows good model performance, with high recall and low FPR at many thresholds. The color gradient shows how performance changes with the probability threshold.

Calculate AUC oos and interpret.

```{r}
#TO-DO
pacman::p_load(pracma)

# Make sure predictions and performance table are ready
asymmetric_predictions_results_oos = as_tibble(
  compute_metrics_prob_classifier(p_hats_test, adult_test$income)
)

# Calculate AUC (Area Under ROC Curve)
auc_oos = -trapz(asymmetric_predictions_results_oos$FPR, asymmetric_predictions_results_oos$recall)
auc_oos
```

#TO-DO interpretation
The out-of-sample AUC is 0.75, which indicates that the model has good discriminatory ability. This means it can correctly rank a randomly chosen positive instance higher than a negative one about 75% of the time.

Plot a FDR-FOR curve oos and interpret.

```{r}
#TO-DO
ggplot(asymmetric_predictions_results_oos) +
  geom_line(aes(x = FDR, y = FOR)) +
  coord_fixed() + 
  xlim(0, max(asymmetric_predictions_results_oos$FDR, na.rm = TRUE)) + 
  ylim(0, max(asymmetric_predictions_results_oos$FOR, na.rm = TRUE)) +
  labs(
    title = "FDR vs. FOR Curve (Out-of-Sample)",
    x = "False Discovery Rate (FDR)",
    y = "False Omission Rate (FOR)"
  )
```

#TO-DO interpretation
This out-of-sample FDR vs. FOR curve shows how the model balances two types of prediction errors. As the false discovery rate (FDR) increases, the false omission rate (FOR) generally decreases, highlighting a trade-off. This helps in selecting a threshold based on which error false positives or false negatives is more costly.

#Boosting

We will make use of YARF's tree-fitting method so here's the boilerplate code to load it once again:

```{r}
options(java.parameters = "-Xmx8000m")
pacman::p_load(rJava)
if (!pacman::p_isinstalled(YARF)){
  pacman::p_install_gh("kapelner/YARF/YARFJARs", ref = "dev")
  pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
}
pacman::p_load(YARF)
```

We will now write a gradient boosting algorithm from scratch. We will make it as general as possible for regression and classification.

```{r}
pacman::p_load(checkmate) #this is a package that enforces arguments are the correct form

#' Gradient boosting
#'
#' Generates a gradient boosting model based on your choices of base learner and objective function
#' 
#' @param X                         A data frame representing the features. It is of size n x p. No need for an intercept column.
#' @param y                         A vector of length n. It either will be real numbers (for regression) or binary (for binary probability estimation).
#' @param g_base_learner_alg        A function with arguments X, y and ... and returns a function that takes X as an argument. The default is YARFCART
#'                                  with nodesize 10% of the total length.
#' @param neg_grad_objective_function   The negative gradient of the function to be minimized. It takes arguments y, yhat that returns a vector. The default objective function is SSE for
#'                                  regression and logistic loss for classification.
#' @param M                         The number of base learners to be summed. Default is 50 for regression and 100 for classification.
#' @param eta                       The step size in the gradient descent. Default is 0.3
#' @param verbose                   Messages are printed out during construction. Default is TRUE.
#' @param ...                       Optional arguments to be passed into the g_base_learner_alg function.
#'
#' @return                          A "qc_basement_gbm" gradient boosting model which can be used for prediction
qc_basement_gbm = function(X, y, g_base_learner_alg = NULL, neg_grad_objective_function = NULL, M = NULL, eta = 0.3, verbose = TRUE, ...){
  assert_data_frame(X)
  n = nrow(X)
  assert_numeric(y)
  assert(length(y) == n)
  assert_function(g_base_learner_alg, args = c("X", "y"), null.ok = TRUE)
  if (is.null(g_base_learner_alg)){
    g_base_learner_alg = function(X0, y0){
      #we want some bias in the base learner - so default to 10% of the sample size
      YARFCART(X0, y0, nodesize = round(.1 * nrow(X0)), calculate_oob_error = FALSE, bootstrap_indices = list(1 : nrow(X0)), verbose = FALSE)
    }
  }
  assert_function(neg_grad_objective_function, args = c("y", "yhat"), null.ok = TRUE)
  assert_count(M, positive = TRUE, null.ok = TRUE)
  assert_numeric(eta, lower = .Machine$double.eps)
  assert_logical(verbose)
  
  if (identical(sort(names(table(y))), c("0", "1"))){
    model_type = "probability_estimation"
    if (verbose){cat("building gradient boosted model for probability estimation of two classes\n")}
    if (is.null(M)){
      M = 100
    }
    if (is.null(neg_grad_objective_function)){
      neg_grad_objective_function = function(y, p_hat){
        y - p_hat
      }
    }
    g_0 = function(X_star){
      rep(mean(y), n)
    }
  } else {
    model_type = "regression"
    if (verbose){cat("building gradient boosted model for regression\n")}
    if (is.null(M)){
      M = 50
    }
    if (is.null(neg_grad_objective_function)){
      neg_grad_objective_function = function(y, y_hat){
        2 * (y - y_hat)
      }
    }
    g_0 = function(X_star){
      rep(mean(y), n)
    }
  }

  g_tildes = list()
  cumul_y_hat_m = g_0(X)
  for (m in 1 : M) {
    if (verbose){cat("fitting base learner", m, "of", M, "\n")}
    g_tildes[[m]] = g_base_learner_alg(X, neg_grad_objective_function(y, cumul_y_hat_m))
    cumul_y_hat_m = cumul_y_hat_m + eta * predict(g_tildes[[m]], X)
  }
  
  gbm = list(
    M = M, 
    eta = eta,
    X = X, 
    y = y, 
    model_type = model_type,
    neg_grad_objective_function = neg_grad_objective_function, 
    g_base_learner_alg = g_base_learner_alg,
    g_0 = g_0,
    g_tildes = g_tildes
  )
  class(gbm) = "qc_basement_gbm"
  gbm
}


#' GBM Predict
#' 
#' Returns final predictions for the gradient boosting model
#'
#' @param gbm     A gradient boosting model of class "qc_basement_gbm"
#' @param X_star  The data to predict for (as a data frame). It has n_* rows and p columns
#'
#' @return        A vector of length n_* rows with each row's predictions. For regression, the
#'                unit is in the units of the original response. For probability estimation for 
#'                binary response, the unit is the logit of the probability estimate.
qc_basement_gbm_predict = function(gbm, X_star){
  y_hat_star = gbm$g_0(X_star)
  for (m in 1 : gbm$M){
    y_hat_star = y_hat_star + gbm$eta * predict(gbm$g_tildes[[m]], X_star)
  }
  y_hat_star
}

#' Compute all iterative boosting predictions
#' 
#' Returns all predictions for each iteration of the gradient boosting
#'
#' @param gbm     A gradient boosting model of class "qc_basement_gbm"
#' @param X_star  The data to predict for (as a data frame). It has n_* rows and p columns
#'
#' @return        A matrix with n_* rows and M+1 columns where each column are the iterative
#'                predictions across all base learners beginning with g_0. For regression, the
#'                unit is in the units of the original response. For probability estimation for 
#'                binary response, the unit is the logit of the probability estimate.
qc_basement_gbm_all_predictions = function(gbm, X_star){
  assert_class(gbm, "qc_basement_gbm")
  assert_data_frame(X_star)
  
  all_y_hat_star = matrix(NA, nrow = nrow(X_star), ncol = gbm$M + 1)
  all_y_hat_star[, 1] = gbm$g_0(X_star)
  for (m in 1 : gbm$M){
    all_y_hat_star[, m + 1] = all_y_hat_star[, m] + 
      gbm$eta * predict(gbm$g_tildes[[m]], X_star)
  } 
  all_y_hat_star
}


#' GBM Predict
#' 
#' Returns final predictions for the gradient boosting model
#'
#' @param gbm     A gradient boosting model of class "qc_basement_gbm"
#' @param X_star  The data to predict for (as a data frame). It has n_* rows and p columns
#'
#' @return        A vector of length n_* rows with each row's predictions. For regression, the
#'                unit is in the units of the original response. For probability estimation for 
#'                binary response, the unit is the logit of the probability estimate.
qc_basement_gbm_predict = function(gbm, X_star){
  qc_basement_gbm_all_predictions(gbm, X_star)[, gbm$M + 1]
}
```

Now we test the code in-sample:

```{r}
set.seed(1)
n = 100
p = 3
X = matrix(rnorm(n * p), nrow = n)
bbeta = seq(-1, 1, length.out = p)
y = c(X %*% bbeta + rnorm(n))
y_binary = rbinom(n, 1, 1 / (1 + exp(-X %*% bbeta)))
X = data.frame(X)

#regression
gbm = qc_basement_gbm(X, y)
pacman::p_load(ggplot2)
ggplot(data.frame(y = y, yhat = qc_basement_gbm_predict(gbm, X))) + 
   geom_point(aes(x = y, y = yhat))
y_hats_by_m = qc_basement_gbm_all_predictions(gbm, X)
rmses_by_m = apply(y_hats_by_m, 2, function(y_hat){sqrt(mean((y - y_hat)^2))})
ggplot(data.frame(m = 1 : (gbm$M + 1), rmse = rmses_by_m)) +
  geom_point(aes(x = m, y = rmse))

#probability estimation
gbm = qc_basement_gbm(X, y_binary)
table(y_binary, as.numeric(qc_basement_gbm_predict(gbm, X) > 0))
y_hats_by_m = qc_basement_gbm_all_predictions(gbm, X) > 0
miscl_err_by_m = apply(y_hats_by_m, 2, function(y_hat){mean(y_binary != y_hat)})
ggplot(data.frame(m = 1 : (gbm$M + 1), miscl_err = miscl_err_by_m)) + 
  geom_point(aes(x = m, y = miscl_err))
```


Here is code to split up the diamonds dataset into three subsets:

```{r}
set.seed(1)
diamonds = ggplot2::diamonds
pacman::p_load(tidyverse)
diamonds = diamonds %>% 
  mutate(cut = factor(cut, ordered = FALSE)) %>%
  mutate(color = factor(color, ordered = FALSE)) %>%
  mutate(clarity = factor(clarity, ordered = FALSE))
diamonds_mm = model.matrix(price ~ ., diamonds)
train_size = 2000
train_indices = sample(1 : nrow(diamonds), train_size)

y_train = diamonds[train_indices, ]$price
X_train = diamonds_mm[train_indices, ]

validation_size = 2000
validation_indices = sample(setdiff(1 : nrow(diamonds), train_indices), validation_size)
y_validation = diamonds[validation_indices, ]$price
X_validation_mm = diamonds_mm[validation_indices, ]

test_size = 2000
test_indices = sample(setdiff(1 : nrow(diamonds), c(train_indices, validation_indices)), test_size)
y_test = diamonds[test_indices, ]$price
X_test_mm = diamonds_mm[test_indices, ]
```

Using your new gradient boosting function, optimize the number of base learners, M for the diamonds data using a grid search:

```{r}
#TO-DO
grid_M = seq(10, 200, by = 10)
validation_rmses = numeric(length(grid_M))

for (i in seq_along(grid_M)) {
  M_val = grid_M[i]
  gbm_model = qc_basement_gbm(data.frame(X_train), y_train, M = M_val, verbose = FALSE)
  yhat_valid = qc_basement_gbm_predict(gbm_model, data.frame(X_validation_mm))
  validation_rmses[i] = sqrt(mean((y_validation - yhat_valid)^2))
}

# Find optimal M
optimal_M = grid_M[which.min(validation_rmses)]
cat("Best M =", optimal_M, "with validation RMSE =", min(validation_rmses), "\n")

# Optional: plot RMSE vs M
pacman::p_load(ggplot2)
ggplot(data.frame(M = grid_M, Validation_RMSE = validation_rmses)) +
  geom_line(aes(x = M, y = Validation_RMSE)) +
  geom_point(aes(x = M, y = Validation_RMSE)) +
  ggtitle("Validation RMSE vs. Number of Boosting Iterations (M)") +
  ylab("Validation RMSE") +
  xlab("Number of Base Learners (M)")
```

Now find the error in the test set and comment on its performance:

```{r}
#TO-DO
# Refit model on training data using optimal M
gbm_best = qc_basement_gbm(data.frame(X_train), y_train, M = optimal_M, verbose = FALSE)

# Predict on test set
yhat_test = qc_basement_gbm_predict(gbm_best, data.frame(X_test_mm))

# Compute test RMSE
test_rmse = sqrt(mean((y_test - yhat_test)^2))
cat("Test RMSE with M =", optimal_M, "is", test_rmse, "\n")
```

Repeat this exercise for the adult dataset. First create the splits:

```{r}
#TO-DO
pacman::p_load_gh("coatless/ucidata")
data(adult)

pacman::p_load(dplyr)
adult = na.omit(adult)
n = nrow(adult)
js = sapply(adult, is.character)
adult[, js] = lapply(adult[, js], as.factor)

# Convert income to binary: 1 = >50K, 0 = <=50K
adult$income = ifelse(adult$income == ">50K", 1, 0)

adult_mm = model.matrix(income ~ ., adult)
y = adult$income

set.seed(1)

# Define sizes
train_size = 2000
validation_size = 2000
test_size = 2000

# Sample indices
train_idx = sample(1:n, train_size)
remaining_idx = setdiff(1:n, train_idx)
validation_idx = sample(remaining_idx, validation_size)
test_idx = sample(setdiff(remaining_idx, validation_idx), test_size)

# Create datasets
X_train = adult_mm[train_idx, ]
y_train = y[train_idx]

X_validation = adult_mm[validation_idx, ]
y_validation = y[validation_idx]

X_test = adult_mm[test_idx, ]
y_test = y[test_idx]
```

Using your new gradient boosting function, optimize the number of base learners, M for the adult data using a grid search:

```{r}
#TO-DO
# Define a grid of M values to search over
grid_M = seq(10, 200, by = 10)
validation_rmses = numeric(length(grid_M))

# Loop through M values
for (i in seq_along(grid_M)) {
  M_val = grid_M[i]
  gbm_model = qc_basement_gbm(data.frame(X_train), y_train, M = M_val, verbose = FALSE)
  yhat_valid = qc_basement_gbm_predict(gbm_model, data.frame(X_validation))
  validation_rmses[i] = sqrt(mean((y_validation - yhat_valid)^2))
}

# Find best M
optimal_M = grid_M[which.min(validation_rmses)]
cat("Best M =", optimal_M, "with validation RMSE =", round(min(validation_rmses), 4), "\n")

library(ggplot2)
ggplot(data.frame(M = grid_M, Validation_RMSE = validation_rmses)) +
  geom_line(aes(x = M, y = Validation_RMSE)) +
  geom_point(aes(x = M, y = Validation_RMSE)) +
  ggtitle("Validation RMSE vs. Number of Base Learners (M)") +
  ylab("Validation RMSE") +
  xlab("Boosting Iterations (M)")
```

Now find the error in the test set and comment on its performance:

```{r}
#TO-DO

# Retrain with best M on training set
gbm_best = qc_basement_gbm(data.frame(X_train), y_train, M = optimal_M, verbose = FALSE)

# Predict on test set
yhat_test_logit = qc_basement_gbm_predict(gbm_best, data.frame(X_test))

# Convert logit to binary prediction using 0 as threshold
yhat_test = as.numeric(yhat_test_logit > 0)

# Misclassification error
test_error = mean(yhat_test != y_test)
cat("Test misclassification error:", round(test_error, 4), "\n")
```


