---
title: "Final_Project"
Name: Ye Htut Maung
date: "2025-05-25"
---

# Packages/Libraries Setup

```{r}
pacman::p_load(rJava, readr, stringr, pander, checkmate, ggplot2, gridExtra, R6, Bolstad2)
if (!pacman::p_isinstalled(YARF)){
  pacman::p_install_gh("kapelner/YARF/YARFJARs", ref = "dev")
  pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
}
pacman::p_load(YARF)
options(java.parameters = "-Xmx8000m")
pacman::p_load(data.table, tidyverse, magrittr, ggplot2, skimr, lubridate, missForest, YARF, rJava)
pacman::p_load(mlr3, mlr3learners, mlr3tuning, mlr3verse, paradox)
```



# Data Import and skimming data

```{r}
Xy = fread("housing_data_2016_2017.csv")
skim(Xy)
head(Xy, 5)
nrow(Xy)
ncol(Xy)
```




# Data Preparation
# dropping irrelevent features
```{r}
# List of columns to drop
cols_to_drop = c(
  "HITId", "HITTypeId", "Title", "Description", "Reward", "CreationTime",
  "RequesterAnnotation", "Expiration", "AssignmentId", "WorkerId",
  "AssignmentStatus", "AcceptTime", "SubmitTime", "AutoApprovalTime",
  "ApprovalTime", "LifetimeApprovalRate", "Last30DaysApprovalRate", "Last7DaysApprovalRate",
  "Keywords", "NumberOfSimilarHITs", "LifetimeInSeconds", "RejectionTime", "RequesterFeedback",
  "MaxAssignments", "AssignmentDurationInSeconds", "AutoApprovalDelayInSeconds", "WorkTimeInSeconds",
  "model_type", "date_of_sale"
) # I drop URL as I can get zipcode from full_address_or_zip_code

# Drop the columns
Xy_new =  Xy[, !names(Xy) %in% cols_to_drop, with = FALSE]
skim(Xy_new)
head(Xy_new, 5)
ncol(Xy_new)
```

We left with 27 featuers.


# cat allowed
``` {r}
# cats_allowed (yes, no) 
Xy_new[["cats_allowed"]]
sum(is.na(Xy_new$cats_allowed)) # no missing
Xy_new[cats_allowed == "y", cats_allowed := "yes"]
Xy_new[, cats_allowed := as.factor(cats_allowed)]

Xy_new[["cats_allowed"]]
summary(Xy_new$cats_allowed)
```


# common_charges and maintenance_cost
``` {r}
# common_charges is for condo
Xy_new[["common_charges"]]
# maintenance_cost is for co op
Xy_new[["maintenance_cost"]]
# combine common_charges and maintenance_cost into one column and called it general_cost
Xy_new[, general_cost := fifelse(!is.na(common_charges), common_charges, maintenance_cost)]
table(Xy_new$general_cost)
#checking
Xy_new[, .(common_charges, maintenance_cost, general_cost)]
# drop the common_charges and maintenance_cost
Xy_new[, c("common_charges", "maintenance_cost") := NULL]
sum(is.na(Xy_new$general_cost)) # 100 missing
# remove $ and , and conver to numeric
Xy_new[, general_cost := as.numeric(gsub("[\\$,]", "", general_cost))]
summary(Xy_new$general_cost)
sd(Xy_new$general_cost, na.rm = TRUE)
```



# coop_condo 
``` {r}

Xy_new[["coop_condo"]]
unique(Xy_new$coop_condo)
sum(is.na(Xy_new$coop_condo)) # no missing

Xy_new[, coop_condo := as.factor(coop_condo)]
table(Xy_new$coop_condo)
```

# dining_room_type (make it factor)
```{r}
table(Xy_new$dining_room_type)
Xy_new[dining_room_type %in% c("dining area", "none"), dining_room_type := "other"]
table(Xy_new$dining_room_type)
Xy_new[["dining_room_type"]]
Xy_new[, dining_room_type := as.factor(dining_room_type)]
levels(Xy_new$dining_room_type)
sum(is.na(Xy_new$dining_room_type))
```

# dogs_allowed
```{r}
table(Xy_new$dogs_allowed)
Xy_new[dogs_allowed == "yes89", dogs_allowed := "yes"]
Xy_new[, dogs_allowed := as.factor(dogs_allowed)]
table(Xy_new$dogs_allowed)
sum(is.na(Xy_new$dogs_allowed))
```

# fuel_type
```{r}
# fuel_type
table(Xy_new$fuel_type)
# Combine "Other" into "other"
Xy_new[fuel_type %in% c("Other"), fuel_type := "other"]
# Convert to factor
Xy_new[, fuel_type := as.factor(fuel_type)]
table(Xy_new$fuel_type)
levels(Xy_new$fuel_type)
sum(is.na(Xy_new$fuel_type))
```

# garage_exists
```{r}

table(Xy_new$garage_exists)
Xy_new[["garage_exists"]]
# Normalize values to lowercase
Xy_new[, garage_exists := tolower(garage_exists)]
# Set valid yes values
Xy_new[garage_exists %in% c("1", "eys", "ug", "underground"), garage_exists := "yes"]
# Create binary indicator: 1 if garage exists, 0 otherwise
Xy_new[is.na(garage_exists), garage_exists := "no"]
Xy_new[, garage_exists := as.factor(garage_exists)]
sum(is.na(Xy_new$garage_exist))
table(Xy_new$garage_exists)
```

# kitchen_type
```{r}
table(Xy_new$kitchen_type)
# Make everything lowercase
Xy_new[, kitchen_type := tolower(kitchen_type)]
# Collapse "eat in" variations into "eat_in"
Xy_new[kitchen_type %in% c("eat in", "eatin"), kitchen_type := "eat_in"]
# Collapse "efficiency" variations into "efficiency"
Xy_new[kitchen_type %in% c("efficiemcy", "efficiency kitchen", "efficiency kitchene", 
                           "efficiency ktchen", "efficiency kitchene"), 
        kitchen_type := "efficiency"]
Xy_new[kitchen_type %in% "1955", kitchen_type := "none"]
# Convert to factor
Xy_new[, kitchen_type := as.factor(kitchen_type)]
# View cleaned result
table(Xy_new$kitchen_type)
sum(is.na(Xy_new$kitchen_type))
```

# parking_charges

```{r}
table(Xy_new$parking_charges)
# Remove $ and convert to numeric
Xy_new[, parking_charges := as.numeric(gsub("\\$", "", parking_charges))]
summary(Xy_new$parking_charges)
sum(is.na(Xy_new$parking_charges))
sd(Xy_new$parking_charges, na.rm=TRUE)
```


# sale_price
```{r}
table(Xy_new$sale_price)
Xy_new[["sale_price"]]
# Remove dollar sign and convert to numeric
Xy_new[, sale_price := as.numeric(gsub("[\\$,]", "", sale_price))]
summary(Xy_new$sale_price)
```

# total_taxes
``` {r}
table(Xy_new$total_taxes)
Xy_new[, total_taxes := as.numeric(gsub("[\\$,]", "", total_taxes))]
summary(Xy_new$total_taxes)
sd(Xy_new$total_taxes, na.rm=TRUE)
```

# listing_price_to_nearest_1000 
```{r}
table(Xy_new$listing_price_to_nearest_1000 )
Xy_new[["listing_price_to_nearest_1000"]]
Xy_new[, listing_price_to_nearest_1000 := as.numeric(gsub("[\\$,]", "", listing_price_to_nearest_1000))]
summary(Xy_new$listing_price_to_nearest_1000)
sd(Xy_new$listing_price_to_nearest_1000, na.rm=TRUE)
```


``` {r}
# approx_year_built
Xy_new[["approx_year_built"]] 
summary(Xy_new$approx_year_built)
sd(Xy_new$approx_year_built, na.rm=TRUE)
```

# community_district_num
```{r}
Xy_new[["community_district_num"]]
table(Xy_new$community_district_num)
Xy_new[, community_district_num := as.character(community_district_num)]
# Define the values to collapse into "other"
rare_districts = c(3, 4, 6, 8, 11, 12, 13, 14, 15, 16, 18, 19, 20, 22, 23, 31, 32)

# Convert those values to "other"
Xy_new[community_district_num %in% rare_districts, community_district_num := "other"]

# Finally, convert to factor
Xy_new[, community_district_num := as.factor(community_district_num)]

table(Xy_new$community_district_num)
summary(Xy_new$community_district_num)
sum(is.na(Xy_new$community_district_num))
```



# num_bedrooms
```{r}
Xy_new[["num_bedrooms"]]
summary(Xy_new$num_bedrooms)
sd(Xy_new$num_bedrooms, na.rm=TRUE)
```


# num_floors_in_building
```{r}
Xy_new[["num_floors_in_building"]]
table(Xy_new$num_floors_in_building)
summary(Xy_new$num_floors_in_building)
sd(Xy_new$num_floors_in_building, na.rm=TRUE)
```

# num_full_bathrooms
``` {r}
Xy_new[["num_full_bathrooms"]]
table(Xy_new$num_full_bathrooms)
summary(Xy_new$num_full_bathrooms)
sd(Xy_new$num_full_bathrooms, na.rm=TRUE)
sum(is.na(Xy_new$num_full_bathrooms))
```


# num_half_bathrooms
```{r}
Xy_new[["num_half_bathrooms"]]
table(Xy_new$num_half_bathrooms)
summary(Xy_new$num_half_bathrooms)
sd(Xy_new$num_half_bathrooms, na.rm=TRUE)
sum(is.na(Xy_new$num_half_bathrooms))
```

# num_total_rooms  
```{r}
Xy_new[["num_total_rooms"]]
table(Xy_new$num_total_rooms)
summary(Xy_new$num_total_rooms)
sd(Xy_new$num_total_rooms, na.rm=TRUE)
```


# pct_tax_deductibl
```{r}
Xy_new[["pct_tax_deductibl"]]
table(Xy_new$pct_tax_deductibl)
summary(Xy_new$pct_tax_deductibl)
sd(Xy_new$pct_tax_deductibl, na.rm=TRUE)
```

# sq_footage 
```{r}
Xy_new[["sq_footage"]]
table(Xy_new$sq_footage)
summary(Xy_new$sq_footage)
sd(Xy_new$sq_footage, na.rm=TRUE)
```


# walk_score
```{r}
Xy_new[["walk_score"]]
table(Xy_new$walk_score)
summary(Xy_new$walk_score)
sd(Xy_new$walk_score, na.rm=TRUE)
```


# full_address_or_zip_code and URL
```{r}

Xy_new[, zip_code := stringr::str_extract(full_address_or_zip_code, "\\d{5}$")]
Xy_new[, "full_address_or_zip_code" := NULL]
# Xy_new[, zip_code := as.factor(zip_code)]


# Create a new column 'final_url' taking non-missing values from URL or url
Xy_new[, final_url := fifelse(!is.na(URL), URL, url)]
Xy_new[, c("URL", "url") := NULL]
head(Xy_new$final_url, 5)
Xy_new[, zip_from_url := stringr::str_extract(final_url, "(?<=-)[0-9]{5}(?=-[0-9]+$)")]

# Xy_new[, zip_from_url := as.factor(zip_from_url)]
# Fill missing zip_code using zip_from_url
Xy_new[is.na(zip_code), zip_code := zip_from_url]
# Drop the temporary zip_from_url column and final_url
Xy_new[,c("final_url", "zip_from_url") := NULL]
table(Xy_new$zip_code)
sum(is.na(Xy_new$zip_code))
# making other for the zip code that is less than 3
# Count frequency of each zip code
zip_counts = table(Xy_new$zip_code)
zip_counts
# Identify zip codes with < 3 instances
rare_zips = names(zip_counts[zip_counts < 3])
rare_zips
# Collapse rare zip codes into "other"
Xy_new[zip_code %in% rare_zips, zip_code := "other"]

# Convert to factor
Xy_new[, zip_code := as.factor(zip_code)]

table(Xy_new$zip_code)
summary(Xy_new$zip_code)
#sd(Xy_new$zip_code, na.rm=TRUE)
```

```{r}
ncol(Xy_new)
```



# Data Cleaning I (Errors, Duplicates, Outliers Handling)

# Data Manipulation (Featurization, Imputation)
```{r}
head(Xy_new, 900)
# ensures non-missing come first (FALSE < TRUE)
Xy_new = Xy_new[order(is.na(sale_price))]
sum(is.na(Xy_new$sale_price))
which(!is.na(Xy_new$sale_price))
last_non_missing_y_index = max(range(which(!is.na(Xy_new$sale_price))))
last_non_missing_y_index
```

# Imputation

```{r}
# Ensure it's a data.frame
is.data.frame(Xy_new)

# Define total number of non-missing y rows
total_non_missing = last_non_missing_y_index

# Set sizes explicitly
select_size = 100
test_size = 100

# Get index ranges
select_indices = 1:select_size
test_indices = (select_size + 1):(select_size + test_size)
train_indices = (select_size + test_size + 1):total_non_missing

# Subset the data
D_select = Xy_new[select_indices]
D_test = Xy_new[test_indices]
D_train = Xy_new[train_indices]

# Extract y values
y_select = D_select$sale_price
y_test = D_test$sale_price
```

# Pre-imputation Locking out the y_test
```{r}
# Start from a copy of Xy_new (already sorted so rows 1:528 have non-missing y)
Xy_pre_impute = copy(Xy_new)

# Combine indices for test and select sets
holdout_indices = c(test_indices, select_indices)

# Mask sale_price values in both sets
Xy_pre_impute[holdout_indices, sale_price := NA]
```

# imputation time

```{r}
# Apply missForest
Xy_imputed = missForest(Xy_pre_impute)$ximp
Xy_imputed
```
# Cleaning after imputation (round numbers that can't have decimals)
```{r}
cols_to_round = c(
  "num_bedrooms", 
  "num_floors_in_building", 
  "num_half_bathrooms", 
  "num_total_rooms", 
  "approx_year_built"
)

# Round and convert to integer
Xy_imputed[, (cols_to_round) := lapply(.SD, function(x) as.integer(round(x))), .SDcols = cols_to_round]
Xy_imputed
```


# spliting into X_train, X_test, y_train, y_test

```{r}
# Training set
X_train = Xy_imputed[train_indices, !"sale_price", with = FALSE]
y_train = Xy_imputed[train_indices, sale_price]

# Test set
X_test = Xy_imputed[test_indices, !"sale_price", with = FALSE]
# y_test was saved before imputation and should not be overwritten

# Select (validation) set
X_select = Xy_imputed[select_indices, !"sale_price", with = FALSE]
# y_select was also saved before imputation and should remain untouched
```



# Regression Tree Modeling
Finding the best nodesize 
```{r}

# Set max nodesize to test
max_nodesize_to_try = 70

# Initialize vectors to store errors
in_sample_errors = numeric(max_nodesize_to_try)
oos_errors = numeric(max_nodesize_to_try)

# Loop through possible nodesize values
for (i in 1:max_nodesize_to_try) {
  tree_mod = YARFCART(data.frame(X_train), y_train, nodesize = i, calculate_oob_error = FALSE)
  
  # In-sample prediction
  yhat_train = predict(tree_mod, data.frame(X_train))
  in_sample_errors[i] = sd(y_train - yhat_train)
  
  # Out-of-sample (validation) prediction
  yhat_select = predict(tree_mod, data.frame(X_select))
  oos_errors[i] = sd(y_select - yhat_select)
}


ggplot(data.frame(
  nodesize = 1:max_nodesize_to_try,
  in_sample_errors = in_sample_errors,
  oos_errors = oos_errors
)) + 
  geom_point(aes(x = nodesize, y = in_sample_errors), col = "red") +
  geom_line(aes(x = nodesize, y = in_sample_errors), col = "red") +
  geom_point(aes(x = nodesize, y = oos_errors), col = "blue") +
  geom_line(aes(x = nodesize, y = oos_errors), col = "blue") +
  labs(
    title = "In-Sample and OOS Error vs Nodesize",
    x = "nodesize",
    y = "Standard Deviation Error"
  ) +
  theme_minimal()
```

```{r}
best_nodesize = which.min(oos_errors)
cat("Best nodesize based on validation set:", best_nodesize, "\n")
```





```{r}

X_train_select = rbind(X_train, X_select)
y_train_select = c(y_train, y_select)
tree_mod = YARFCART(X_train_select,y_train_select, nodesize = best_nodesize, calculate_oob_error = FALSE)
```
# illustrate the tree
```{r}
get_tree_num_nodes_leaves_max_depths(tree_mod)
illustrate_trees(tree_mod, max_depth = 6, open_file = TRUE)

```

# prediction on Regression tree 
# in-sample
```{r}

y_hat_train_select = predict(tree_mod, X_train_select)
rmse_train_select = sqrt(mean((y_train_select - y_hat_train_select)^2))
ss_res_train_select = sum((y_train_select - y_hat_train_select)^2)
ss_tot_train_select = sum((y_train_select - mean(y_train_select))^2)
r2_train_select = 1 - ss_res_train_select / ss_tot_train_select
rmse_train_select
r2_train_select
```



# out-of-sample
```{r}
y_hat_test = predict(tree_mod, X_test)
oos_rmse_tree = sqrt(mean((y_test - y_hat_test)^2))
ss_res = sum((y_test - y_hat_test)^2)
ss_tot = sum((y_test - mean(y_test))^2)
oos_r2_tree = 1 - ss_res / ss_tot
oos_rmse_tree
oos_r2_tree
```



# Linear Modeling

```{r}
lm_model = lm(y_train_select ~ ., data = data.frame(X_train_select, y_train_select))
round(coef(lm_model), 3)

```

# in sample residual
```{r}
y_hat_lm_train_select = predict(lm_model,X_train_select)
rmse_lm_train_select = sqrt(mean((y_train_select - y_hat_lm_train_select)^2))
rmse_lm_train_select

ss_res = sum((y_train_select - y_hat_lm_train_select)^2)
ss_tot = sum((y_train_select - mean(y_train_select))^2)

r2_lm_train_select = 1 - ss_res / ss_tot
r2_lm_train_select
```

# oos
```{r}
oos_y_hat_lm_test = predict(lm_model, X_test)
oos_rmse_lm_test = sqrt(mean((y_test - oos_y_hat_lm_test)^2))
oos_rmse_lm_test
ss_res = sum((y_test - oos_y_hat_lm_test)^2)
ss_tot = sum((y_test - mean(y_test))^2)

r2_lm_test = 1 - ss_res / ss_tot
r2_lm_test
```




# Random Forest Modeling

```{r}
library(randomForest)
# Combine X_train and y_train into a single data frame
df_train = data.frame(X_train)
df_train$sale_price = y_train

df_select = data.frame(X_select)
df_select$sale_price = y_select

p = ncol(X_train)  # number of predictors
num_trees = 1200
oos_se_by_mtry = numeric(p)

# Loop over mtry from 1 to p
for (m in 1:p) {
  rf_model = randomForest(sale_price ~ ., data = df_train, ntree = num_trees, mtry = m)
  y_hat_select = predict(rf_model, newdata = df_select)
  oos_se_by_mtry[m] = sqrt(mean((df_select$sale_price - y_hat_select)^2))  # RMSE
}

# Create data frame for plotting
df_mtry = data.frame(
  mtry = 1:p,
  oos_rmse = oos_se_by_mtry
)

# Plot OOS RMSE vs mtry
library(ggplot2)
ggplot(df_mtry, aes(x = mtry, y = oos_rmse)) +
  geom_point() +
  geom_line() +
  labs(
    title = "OOS RMSE vs mtry (500 Trees)",
    x = "mtry (Number of Features Tried at Each Split)",
    y = "Out-of-Sample RMSE on Select Set"
  ) +
  theme_minimal()

# Find best mtry and fit final model
best_mtry = which.min(oos_se_by_mtry)
cat("Best mtry based on OOS RMSE:", best_mtry, "\n")

```


```{r}
seed = 42
num_trees = 1000
mod_rf = YARF(X_train_select, y_train_select, num_trees = num_trees, seed = seed, mtry = best_mtry, nodesize = 1,    calculate_oob_error = FALSE, verbose = FALSE)
mod_rf
```
# in-sample
```{r}

y_hat_rf_train_select = predict(mod_rf, X_train_select)

rmse_rf_train_select = sqrt(mean((y_train_select - y_hat_rf_train_select)^2))

ss_res_train_select = sum((y_train_select - y_hat_rf_train_select)^2)
ss_tot_train_select = sum((y_train_select - mean(y_train_select))^2)
r2_rf_train_select = 1 - ss_res_train_select / ss_tot_train_select

cat("In-sample RMSE:", rmse_rf_train_select, "\n")
cat("In-sample R²:", r2_rf_train_select, "\n")
```


# out-of-sample 
```{r}

y_hat_rf_test = predict(mod_rf, X_test)
rmse_rf_test = sqrt(mean((y_test - y_hat_rf_test)^2))
ss_res_test = sum((y_test - y_hat_rf_test)^2)
ss_tot_test = sum((y_test - mean(y_test))^2)
r2_rf_test = 1 - ss_res_test / ss_tot_test
cat("Out-of-sample RMSE:", rmse_rf_test, "\n")
cat("Out-of-sample R²:", r2_rf_test, "\n")
```


# Shipping Final Model Trained On All Data
```{r}
# Combine X and y from train, select, and test
X_final = rbind(X_train, X_select, X_test)
y_final = c(y_train, y_select, y_test)

# Fit final YARF model
mod_rf_final = YARF(X_final, y_final, 
                    num_trees = num_trees, 
                    seed = seed, 
                    mtry = best_mtry,
                    nodesize = 1,   calculate_oob_error = FALSE, verbose = FALSE)
mod_rf_final
```


```{r}

```


